### 2013-10-01 src2/expressive/discussion.tex
apa

\section {Discussion}

@@

The series of kernels also illustrates how the use of combinators brings a form of reuse, and makes design exploration easier. Our experience of using similar combinators in the Lava hardware
description language~\cite{LavaSorter} was that a relatively small set of combinators went a long way. So, although we introduced three
combinators here, {\tt ilv}, {\tt vee} and {\tt ilvVee}, which
includes the other two, we do not believe that every new kernel development exercise would demand a completely new set of combinators.
We expect to provide the user with a well-documented set of combinators,
so that users can get access to this style of programming without having
to develop their own combinators, and without having to think too much
about bit-hacking.
The bit-manipulation approach chosen to define our combinators automatically
created functions that apply to sub-sequences of the input that
are of an appropriate length.

>>>>

\subsection {Discussion}

@@

The series of kernels also illustrates how the use of combinators brings a form of reuse, and makes design exploration easier. Our experience of using similar combinators in the Lava hardware
description language~\cite{LAVASORTER} was that a relatively small set of combinators went a long way. So, although we introduced three
combinators here, {\tt ilv}, {\tt vee} and {\tt ilvVee}, which
includes the other two, we do not believe that every new kernel development exercise would demand a completely new set of combinators.
We expect to provide the user with a well-documented set of combinators,
so that users can get access to this style of programming without having
to develop their own combinators, and without having to think too much
about bit-hacking.
The bit-manipulation approach chosen to define our combinators automatically
created functions that apply to sub-sequences of the input that
are of an appropriate length.

### 2013-10-01 src2/expressive/intro.tex
apa

Our approach is different from that of other Haskell DSELs targetting
GPUs~\cite{ACCELERATE,NIKOLA,BARRACUDA}. 
We do not try to abstract away from 
all the peculiarities of GPU programming, but rather provide a higher 
level language in which to experiment with them.
For instance,
Accelerate provides a standard set of basic operations such as 
{\tt map}, {\tt reduce} and {\tt zipWith} as built in skeletons, implemented
with the help of small, predefined, hand-tuned CUDA kernels~\cite{ACCELERATE}. 
Obsidian, on the other hand, allows the user to experiment with the
{\em generation} of small kernels for fixed size array inputs
from higher level descriptions.
It is intended to allow the user to play with the kinds of tradeoffs that are important
when writing such high performance building blocks; in this paper, the main consideration
is the number of array elements of the input and output that are manipulated
by a single thread in the generated CUDA code.
An important aspect of Obsidian is the symbolic array representation used, along
with its associated {\tt sync} operation. As we shall see, the {\tt sync} operation
allows the programmer to guide code generation and control parallelism and thread
use~\cite{JSLIC}.

>>>>

Our approach is different from that of other Haskell DSELs targetting
GPUs~\cite{ACCELERATEDAMP11,NIKOLA,BARRACUDA}. 
We do not try to abstract away from 
all the peculiarities of GPU programming, but rather provide a higher 
level language in which to experiment with them.
For instance,
Accelerate provides a standard set of basic operations such as 
{\tt map}, {\tt reduce} and {\tt zipWith} as built in skeletons, implemented
with the help of small, predefined, hand-tuned CUDA kernels~\cite{ACCELERATEDAMP11}. 
Obsidian, on the other hand, allows the user to experiment with the
{\em generation} of small kernels for fixed size array inputs
from higher level descriptions.
It is intended to allow the user to play with the kinds of tradeoffs that are important
when writing such high performance building blocks; in this paper, the main consideration
is the number of array elements of the input and output that are manipulated
by a single thread in the generated CUDA code.
An important aspect of Obsidian is the symbolic array representation used, along
with its associated {\tt sync} operation. As we shall see, the {\tt sync} operation
allows the programmer to guide code generation and control parallelism and thread
use~\cite{JSLIC}.

### 2013-10-01 src2/ifl/related.tex
apa

Microsoft's Accelerator project moves even closer to general purpose programming by doing away with the kernel notion and simply expressing programs in a data parallel style, using functions on arrays~\cite{Accelerator}.
Data Parallel Haskell~\cite{DPH} incorporates Nested Data Parallelism in the style of NESL~\cite{NESL} into Haskell. GPUGen, like Obsidian, aims
to support GPGPU programming from Haskell~\cite{GPUGen}. It works by translating Haskell's intermediate language, Core, into CUDA, for collective data operations such as scan, fold and map. The intention is to plug GPUGen into the Nested Data Parallel framework of the Glasgow Haskell Compiler. Our impression is that we wish to expose considerably more detail about the GPU to the programmer, but we do not yet have sufficient information about GPUGen to be able to do a more complete comparison. Finally, we mention the Spiral project, which develops methods and tools for automatically generating high performance libraries for a variety of platforms, in domains such a signal processing, multiplication and sorting~\cite{Spiral}.
The tuning of an algorithm for a given platform is expressed as an optimisation problem, and the domain specific mathematical structure of the algorithm is used
to create a feedback-driven optimiser. The results are indeed impressive, and we feel that the approach based on an algebra of what we would call combinators
will interest functional programmers. We hope to experiment with similar
search and learning based methods, having applied similar ideas in the simpler setting of arithmetic data-path generation in Lava.

>>>>

Microsoft's Accelerator project moves even closer to general purpose programming by doing away with the kernel notion and simply expressing programs in a data parallel style, using functions on arrays~\cite{ACCELERATOR}.
Data Parallel Haskell~\cite{DPH} incorporates Nested Data Parallelism in the style of NESL~\cite{NESL} into Haskell. GPUGen, like Obsidian, aims
to support GPGPU programming from Haskell~\cite{GPUGenTalk}. It works by translating Haskell's intermediate language, Core, into CUDA, for collective data operations such as scan, fold and map. The intention is to plug GPUGen into the Nested Data Parallel framework of the Glasgow Haskell Compiler. Our impression is that we wish to expose considerably more detail about the GPU to the programmer, but we do not yet have sufficient information about GPUGen to be able to do a more complete comparison. Finally, we mention the Spiral project, which develops methods and tools for automatically generating high performance libraries for a variety of platforms, in domains such a signal processing, multiplication and sorting~\cite{Spiral}.
The tuning of an algorithm for a given platform is expressed as an optimisation problem, and the domain specific mathematical structure of the algorithm is used
to create a feedback-driven optimiser. The results are indeed impressive, and we feel that the approach based on an algebra of what we would call combinators
will interest functional programmers. We hope to experiment with similar
search and learning based methods, having applied similar ideas in the simpler setting of arithmetic data-path generation in Lava.

### 2013-10-01 src2/ifl/results.tex
apa

\section{Results} \label{sec:results}

Apart from the parallel prefix algorithm shown in this paper we have used
Obsidian to implement sorters. For the sorters, the generated C code 
performs quite well. One periodic sorting network, called {\em Vsort}, implemented in Obsidian in the style of the sorters presented in reference~\cite{LavaSorter},
has a running time of 95$\mu$secs for 256 elements (the largest size that we can cope with in a single kernel). This running time can be compared to the 
 28$\mu$secs running time of the bitonic sort example supplied with the CUDA 
SDK. However, for 256 inputs, bitonic sort has a depth (counted in number of comparators between input and output), of 36 compared to 
Vsort's 64. So  we feel confident that we can make a sorter that improves considerably on our Vsort by implementing a recursive algorithm for which the corresponding network depth is less.
(We implemented the periodic sorter in an earlier version of the system, in which recursion was not available.) The point here is not to be as fast as hand-crafted library code, but to come close enough to allow the user to quickly construct short readable programs that give decent performance. The results for sorting are promising in this respect.
Sadly, the results for the Sklansky example are rather poor, and we will return to this point in the following section.

>>>>

\subsection{Results} \label{sec:results}

Apart from the parallel prefix algorithm shown in this paper we have used
Obsidian to implement sorters. For the sorters, the generated C code 
performs quite well. One periodic sorting network, called {\em Vsort}, implemented in Obsidian in the style of the sorters presented in reference~\cite{LAVASORTER},
has a running time of 95$\mu$secs for 256 elements (the largest size that we can cope with in a single kernel). This running time can be compared to the 
 28$\mu$secs running time of the bitonic sort example supplied with the CUDA 
SDK. However, for 256 inputs, bitonic sort has a depth (counted in number of comparators between input and output), of 36 compared to 
Vsort's 64. So  we feel confident that we can make a sorter that improves considerably on our Vsort by implementing a recursive algorithm for which the corresponding network depth is less.
(We implemented the periodic sorter in an earlier version of the system, in which recursion was not available.) The point here is not to be as fast as hand-crafted library code, but to come close enough to allow the user to quickly construct short readable programs that give decent performance. The results for sorting are promising in this respect.
Sadly, the results for the Sklansky example are rather poor, and we will return to this point in the following section.

### 2013-10-01 src2/papp/paperThesis.tex
apa

Sorting is a popular function to place on the GPU, see~\cite{SORT1,SORT3}.
{\em Odd-Even merge sort}, due to Batcher, uses a merger component called the Odd-Even merger. This 
merger merges two sorted arrays. In Obsidian this merger and sorter 
can be implemented using the combinators {\tt two}, {\tt ilv}, {\tt evens}
and {\tt odds}. For a detailed description of these combinators see
~\cite{lavaSorter}.

@@

This subsection shows the implementation of a parallel prefix (also called scan) 
kernel, known as {\tt sklansky} after J. Sklansky \cite{Sklansky:1960:CSA}.
This kernel will then be optimised step-by-step using Obsidian. This 
optimisation effort uses an experimental feature of Obsidian called {\tt syncHow}. 
This is different from the normal {\tt sync} in the way it assigns work to threads. 
For a more thorough explanation of {\tt syncHow} see ~\cite{JSTECH}.
For an explanation of the parallel-prefix operation see~\cite{BlellochTR90}.

@@

%-------------------------------------------------------------------------------
\FloatBarrier
\subsection{Future work}
\label{sec:FutureWork}

@@

Vertigo is another GPU programming language embedded in Haskell\cite{Vertigo}.
Vertigo can be used to describe procedureal surfaces and textures and generates
efficient DirectX9 shader code.

There are also many examples of domain specific languages (DSLs)
that target not only GPUs but also other platforms, such as multicore machines.
Of particular interest to us is Microsoft's Accelerator project~\cite{Accelerator}. Interestingly, Accelerator includes the same restriction to ``hardware-like'' algorithms as Obsidian does.
%However, Vertigo is targeting graphical applications.

>>>>

Sorting is a popular function to place on the GPU, see~\cite{SORT1,SORT3}.
{\em Odd-Even merge sort}, due to Batcher, uses a merger component called the Odd-Even merger. This 
merger merges two sorted arrays. In Obsidian this merger and sorter 
can be implemented using the combinators {\tt two}, {\tt ilv}, {\tt evens}
and {\tt odds}. For a detailed description of these combinators see
~\cite{LAVASORTER}.

@@

This subsection shows the implementation of a parallel prefix (also called scan) 
kernel, known as {\tt sklansky} after J. Sklansky \cite{Sklansky}.
This kernel will then be optimised step-by-step using Obsidian. This 
optimisation effort uses an experimental feature of Obsidian called {\tt syncHow}. 
This is different from the normal {\tt sync} in the way it assigns work to threads. 
For a more thorough explanation of {\tt syncHow} see ~\cite{JSTECH}.
For an explanation of the parallel-prefix operation see~\cite{BlellochTR90}.

@@

%-------------------------------------------------------------------------------
\FloatBarrier
\subsection{Future work}
% \label{sec:PAPPFutureWork}

@@

Vertigo is another GPU programming language embedded in Haskell\cite{VERTIGO}.
Vertigo can be used to describe procedureal surfaces and textures and generates
efficient DirectX9 shader code.

There are also many examples of domain specific languages (DSLs)
that target not only GPUs but also other platforms, such as multicore machines.
Of particular interest to us is Microsoft's Accelerator project~\cite{ACCELERATOR}. Interestingly, Accelerator includes the same restriction to ``hardware-like'' algorithms as Obsidian does.
%However, Vertigo is targeting graphical applications.
